---
title: "Mothers at Risk"
author: "Dr. Isuru Dassanayake"
output:
  html_document: default
---

```{r, message=FALSE}
library(tidyverse)
library(gridExtra)
```

```{r,echo=FALSE}
hw <- theme_gray()+ theme(
  plot.title=element_text(hjust=0.5),
  plot.subtitle=element_text(hjust=0.5),
  plot.caption=element_text(hjust=-.5),

  strip.text.y = element_blank(),
  strip.background=element_rect(fill=rgb(.9,.95,1),
    colour=gray(.5), size=.2),

  panel.border=element_rect(fill=FALSE,colour=gray(.70)),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank(),
  panel.spacing.x = unit(0.10,"cm"),
  panel.spacing.y = unit(0.05,"cm"),

# axis.ticks.y= element_blank()
  axis.ticks=element_blank(),
  axis.text=element_text(colour="black"),
  axis.text.y=element_text(margin=margin(0,3,0,3)),
  axis.text.x=element_text(margin=margin(-1,0,3,0))
)
```

```{r,echo=FALSE}
hwLeft <- theme_gray()+ theme(
  plot.title=element_blank(),
  plot.subtitle=element_text(hjust=0.5),
  plot.caption=element_text(hjust=-.5),

  strip.text.y = element_blank(),
  strip.background=element_rect(fill=rgb(.9,.95,1),
    colour=gray(.5), size=.2),

  panel.border=element_rect(fill=FALSE,colour=gray(.70)),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank(),
  panel.spacing.x = unit(0.10,"cm"),
  panel.spacing.y = unit(0.05,"cm"),

# axis.ticks.y= element_blank()
  axis.ticks=element_blank(),
  axis.text=element_text(colour="black"),
  axis.text.y=element_text(margin=margin(0,3,0,3)),
  axis.text.x=element_text(margin=margin(-1,0,3,0))
)
```

```{r, echo=FALSE}

hwRight <- theme_gray()+ theme(
  plot.title=element_blank(),
  plot.subtitle=element_text(hjust=0.5),
  plot.caption=element_text(hjust=-.5),

  strip.text.y = element_blank(),
  strip.background=element_rect(fill=rgb(.9,.95,1),
    colour=gray(.5), size=.2),

  panel.border=element_rect(fill=FALSE,colour=gray(.70)),
  panel.grid.minor.y = element_blank(),
  panel.grid.minor.x = element_blank(),
  panel.spacing.x = unit(0.10,"cm"),
  panel.spacing.y = unit(0.05,"cm"),

# axis.ticks.y= element_blank()
  axis.ticks=element_blank(),
  axis.text=element_text(colour="black"),
  axis.text.y=element_blank(),
  axis.text.x=element_text(margin=margin(-1,0,3,0))
)
```

&nbsp;

#### 1. Enter data as vectors
```{r}
attend <- c(25,8,12,31,2,17,15,30,8,8,6,16,24,26,24,15,47,38,44,20,31)
deaths <- c(18,17,16,16,16,15,15,15,15,14,14,14,13,13,12,12,12,12,11,10,10)
nams <- c('Sierra Leone','Afghanistan','Bhutan','Guinea','Somalia','Angola','Chad',
          'Mozambique','Nepal','Ethiopia','Eritrea','Yemen','Burundi','Rwanda',
          'Mali','Niger','Senegal','Uganda','Gambia','Haiti','Nigeria')
```

&nbsp;

#### 2. Make a tibble
```{r}
motherDeaths = tibble(Countries = nams,
  DeathRates = deaths, Attended = attend)

#write.csv(motherDeaths,file = "motherDeaths.csv",
#  row.names = FALSE)

```

&nbsp;

#### 3. Make a scatterplot with a smooth
```{r}
plt <-  ggplot(motherDeaths,
  aes(x = Attended, y = DeathRates)) +
  geom_smooth(method="loess",span=.90,method.args=list(degree=1),
      size=1.5,color="blue") +
  geom_point(shape=21,size=4,color="black",fill="red") +
  labs(x="Percent Births Attended By Trained Personnel",
       y="Maternal Deaths Per 1000 Live Births",
       title="Maternal Risk 1990-1996") + hw
plt
```

&nbsp;

#### 4. Select Countries and Add labels to Points

Below %>% pipes the motherDeaths data.frame into the first argument of filter(). The filter function selects rows of the data.frame.
```{r}
ptLabs <- motherDeaths %>% filter(Countries %in%
     c('Sierra Leone','Guinea','Haiti','Nigeria'))
ptLabs
```

Below we use the selected Country names as labels. We have nudged the labels upward by 45 y-axis units.

```{r}
plt + geom_label(data = ptLabs,
      aes(label = Countries),
      nudge_y = .9)
```


&nbsp;

#### 5. Make row labeled plot with juxtaposed columns for two different variables 


The ggplot facet_grid and facet_wrap are often useful for producing multiple panel layouts. We will make frequent use of this latter.

The script below shows an alternative way to produce multiple panel plots. It uses grid.arrange() from the gridExtra package.

We save the plots as named objects and put them the panels. We can control the number rows and columns of panels.  As as far as I know we cannot control the horizontal spacing between the panels using grid.arrange Likely this can be done with ggtable.

&nbsp;

##### 5.1 Test the left column dot plot

```{r}
ggplot(motherDeaths,aes(x = Attended,
  y = reorder(Countries,-Attended))) +
  geom_point(shape = 21,fill = "cyan",
    color = 'black',size = 3) +
  labs(x = "Percent Births Attended\nBy Trained Personnel",
     y = "Countries") +  hw

```

&nbsp;

##### 5.2 Make the Left column Dot Plot with the Country Names

Omit the title and the y axis label use the theme in hwLeft

```{r}
pltLeft <- ggplot(motherDeaths,aes(x = Attended,
  y = reorder(Countries,-Attended))) +
  geom_point(shape = 21,fill = "cyan",
    color = 'black',size = 3.5) +
  labs(y = "",
     x = "Percent Births\n Attended By Trained Personnel") +
  hwLeft

pltLeft
```

&nbsp;

##### 5.3 Make the Right column Dot Plots without the Country names 

Omit the title
```{r}
pltRight <- ggplot(motherDeaths,aes(x = DeathRates,
  y = reorder(Countries,-Attended))) +
  geom_point(shape = 21,fill = "red",
    color = 'black',size = 3.5) +
  labs(y = "",
     x = "Maternal Deaths\nPer 1000 Live Births") +  hwRight
pltRight

```

&nbsp;

##### 5.4 put the two together and add a title using grid.arrange

```{r}
grid.arrange(pltLeft,pltRight,ncol = 2,widths = c(2.6,2),
  top = "Maternal Risk Giving Birth\n For Selected Countries 1990-1965")
```

&nbsp;

#### 6. Creating Perceptual Groups of size 3

Here we modify the data.frame and give it a new name.

The construction:

1. Arranges the data.frame rows so the countries are in the top down order that we want in the plot.
2. Change the country factor level to be in the reverse order of the top down order consistent with plot y-axis convention.
3. Adds a row perceptual row grouping factorwith levels the keep groups of rows in the same top down order.

&nbsp;

##### 6.1 Sort the data frame rows so the Country order match that in 5. above Revise the factor level match.

##### Get the topDown order
```{r}
newLevels <- levels( with(motherDeaths,reorder(Countries,-Attended)))
topDown <- rev(newLevels)
topDown
```



##### Rearrange rows
```{r}
subs <- match(topDown,motherDeaths$Countries)
mdSort <- motherDeaths[subs,]  # rearrange rows
```



##### Make factor levels provide the top down order
```{r}
mdSort$Countries <- factor(topDown,rev(topDown))

 #or
# Sort the rows in the desired order

ord <- order(motherDeaths$Attended)
mdSort <- motherDeaths[ord,]
mdSort

```


&nbsp;

##### 6.2  Add a grouping factor

We have 21 rows. Here we use 7 groups (labeled G1 to G7) of size 3. 
 
Another reasonable choice is a 5-5-1-5-5 pattern the facet grid argument space="free_y" will give less space to the group of size 1.

```{r}
nam <- paste0("G",7:1)
Grp <- factor(rep(nam,each = 3),levels=nam)
mdSort$Grp <- Grp


pLeft <- ggplot(mdSort,aes(x = Attended, y = Countries)) +
  geom_point(shape = 21,fill = "cyan",color = 'black',size = 2.5)+
  facet_grid(Grp~.,scales = "free_y") +
  labs(y = "",
     x = "Percent Births\n Attended By Trained Personnel") + hwLeft
pLeft

pRight <- ggplot(mdSort,aes(x = DeathRates, y = Countries)) +
  geom_point(shape = 21,fill = "red",color = 'black',size = 2.5) +
  facet_grid(Grp~.,scales = "free_y") +
  labs(y="", x="Maternal Deaths\nPer 1000 Live Births") + hwRight
pRight

grid.arrange(pLeft,pRight,ncol=2,widths=c(2.6,2),
  top="Maternal Risk Giving Birth\n For Selected Countries 1990-1965")
```






